\chameleon{}: Untangling Type Errors Through Interactive Visualization and Exploration

### 2022-12-08 sections/discussion.tex
More Tim's edits

\todo{Suggest try and strucuture this a bit more e.g. each para bold the key lesson @ start of para. Can you group into implications for research/implications for practice to make more actionable??}

This paper presents the interactive type debugging tool \chameleon{} and charts the evolution of its design across several iterations in response to user evaluation and feedback, as well as examining the effectiveness of the general approach compared to traditional static type error messages. We found that programmers using \chameleon{} type compare tool are able to debug errors faster than using traditional text-based error messages. This effect is shown more clearly when the task is not trivial. We found that programmers who actively use ChameleonIDE's interactive features (candidate expression cards and deduction steps) are more efficient in fixing type errors than simply reading the type error output. In this section, we will discuss a few interpretations of the results.

From the results of user studies 1 and 2, we observed that the choice of debugging tool had little effect on how fast programmers solve type errors. Conversely, when facing more realistic problems (longer source code, error locations more scattered), programmers are more effective using \chameleon{}. We believe one factor of this effectiveness of \chameleon{} is to reduce the amount of reading time. Jbara \cite{jbara_how_2015} showed that reading source code is generally the initial step of solving programming problems and is done in several passes. Although traditional compiler error message tools initially show fewer locations, it is unreliable, meaning that programmers have to expand the reading span without clear guidance. In contrast, Chameleon shows more error locations initially. However, the completeness of error locations assures programmers which part of the source code can be safely skipped.

From the results of user studies 3 and 4, we found that programmers who use interactive tool fix type errors faster than the ones who passively read the error output. This difference is stronger in harder tasks. We speculate that one factor of this result is that  \chameleon{} helps to develop debugging plans. We observed that when working with \chameleon{}, programmers form different debugging tactics to attack the problem. Among the high interactivity participants in user study 4, some programmers cycle through deduction steps as a guide to reading source code; some programmers navigate to both ends of the deduction steps where types are normally grounded and concrete. In contrast, minimal interactive participants generally form more similar plans, including carefully reading the program text and manually annotating expressions based on their understanding of the program.

\begin{listing}[!ht]
\begin{minted}{haskell}
f z     
    | z == 3 = False
    | z == '4' = True
\end{minted}
\caption{     In this simple program, \chameleon{} report the error happened in \texttt{f} and \texttt{z}.}
\label{listing:2}
\end{listing}

We speculate another factor of the effectiveness \chameleon{} interactive debugging tools is they help programmers effectively chunk intermediate information. With the program showing in the listing \ref{listing:2}, \chameleon{} offers two candidate expressions: \texttt{f} can be typed as \texttt{Int -> Bool} or \texttt{Char -> Bool}; \texttt{z} can be typed as \texttt{Int} or \texttt{Char}. Although, in theory, these two statements are to the same effect, programmers are often required to compute the latter from the former or vice versa. And this computation may carry out multiple layers. Programmers have to remember all the intermediate types and their reasoning throughout such mental gymnastics. Assisted by candidate expression cards and deduction steps, this intermediate information is externalized on screen and can be retrieved anytime. A recent study on working memory \cite{crichton_role_2021} suggested this approach may provide a positive effect in helping programmers manage cognitive load and free up working-memory space for high-level thinking.

In our studies, \chameleon{} shows a few different designs for type error visualization and interaction. However, we are far from exhausting the search space. One challenge we notice in the current design is that programmers have to shift their focus between the editor on the left and the \chameleon{} debugging interface on the right. Using the hover popup window in most mainstream IDEs may reduce the context switching during debugging.

On a similar note, the current implementation of \chameleon{} requires non-trivial adaptation for editors such as VS Code and IntelliJ due to the overlay explanation layer. This type of error visualization is non-standard in mainstream editors. However, there may exist alternative representations of \chameleon{} error reporting using only the features available in major editors and IDEs. It is especially beneficial to represent \chameleon{} errors using a universal debugging middle layer such as language server protocol (LSP). This will allow \chameleon{}  to be adapted into various coding environments which support an LSP back end.

Another area for future study is to extend the evaluation process. In studies 1b and 2, we used harder tasks compared to study 1a. However, the difficulty level is nowhere as hard as a professional Haskell programmer may face in the production Haskell codebase. It would be interesting to see how \chameleon{}  faces type errors that span multiple files and packages. Another dimension of difficulty is higher-level extractions. Haskell programs get more tricky to the debugger when more abstractions are added to the codebase, such as Monads, Monad transformers, and Lenses. These topics were not included in our study to accommodate novice users. However, observing how expert Haskell programmers use \chameleon{} features to solve advanced problems will gain valuable insight into the true usability of the tool.

Another direction to expand the work is in data collection. The online study limited the amount of qualitative data we could gather from the users. Although we handed out debriefing surveys after each study, it is no substitute for more in-depth qualitative study methods. Inviting users for speak-aloud walkthroughs and interviews are viable methods to unveil deeper usability issues about \chameleon{}.

Our study focuses on the Haskell language for its popularity in the academic world. However, the low quality of textual type errors is a problem not limited to Haskell. Modern statically typed languages more or less all share the same problem. We believe the underlying type reporting technique and algorithms can be generalized to other languages. It will be exciting to see how interactive debugging features perform in other paradigms, such as  imperative languages and incremental type systems.

Our design of \chameleon{}  emphasizes allowing programmers to update the type error based on their prior knowledge. It unfolds to answer three questions: which expression should be considered un-typeable? Between multiple possible types, which one should be considered intended? Among all the possible locations, which one should be considered the cause? \chameleon{} answers all three questions by leaving it to the programmers to decide. However, it is possible and likely that some programmers enjoy conclusive error messages instead of flexible ones provided by \chameleon{}. Using heuristic methods may achieve the best of both worlds. It may produce effective results by guessing the possible un-typable expression based on the position it appears in the syntax tree and project structure or guessing the intended type signature by comparing the number of supporting slices in the source code.

>>>>

% \todo{Suggest try and strucuture this a bit more e.g. each para bold the key lesson @ start of para. Can you group into implications for research/implications for practice to make more actionable??}

This paper presents the interactive type debugging tool \chameleon{} and charts the evolution of its design across several iterations in response to user evaluation and feedback, as well as examining the effectiveness of the general approach compared to traditional static type error messages. We found that programmers using \chameleon{} type compare tool are able to debug errors faster than using traditional text-based error messages. This effect is shown more clearly when the task is not trivial. We found that programmers who actively use \chameleon{}'s interactive features (candidate expression cards and deduction steps) are more efficient in fixing type errors than passively reading the type error output. In this section, we will discuss a few interpretations of the results.

\subsection{Effect on Reading Source Code}
From the results of Study 1a, we observed that the choice of debugging tool had little effect on how fast programmers solve simple type errors. Conversely, when facing more realistic problems (longer source code, error locations more scattered) in study 1b, programmers are more effective using \chameleon{}. One explanation is that \chameleon{} reduces the amount of reading time by taking programmers more directly to the problem. Earlier studies \cite{jbara_how_2015, peitek_what_2020} showed that reading source code is generally the initial step of solving programming problems and is done in several passes. Although traditional compiler error message tools initially show fewer locations, these may be incomplete, meaning that programmers have to expand the reading span without clear guidance. In contrast, Chameleon shows more error locations initially. However, the completeness of error locations assures programmers which part of the source code can be safely skipped.

\subsection{Forming Debugging Plans}
From the results of Study 2, we found that programmers who use the interactive tool fix type errors faster than the ones who passively read the error output. This effect is stronger in harder tasks. We speculate that one factor of this result is that  \chameleon{} helps to develop debugging plans. We observed that when working with \chameleon{}, programmers form different debugging tactics to attack the problem. Among the \textit{high} interactivity participants in user study 4, some programmers cycle through deduction steps as a guide to reading source code; some programmers navigate to both ends of the deduction steps where types are normally grounded and concrete. In contrast, \textit{minimal} interactive participants generally form similar plans, including carefully reading the program text and manually annotating expressions based on their understanding of the program.

\subsection{Externalise Intermediate Typing Inforamtion}
We speculate another factor of the effectiveness \chameleon{} interactive debugging tools is they help programmers effectively chunk intermediate information. With the program showing in the listing \ref{listing:2}, \chameleon{} offers two candidate expressions: \texttt{f} can be typed as \texttt{Int -> Bool} or \texttt{Char -> Bool}; \texttt{z} can be typed as \texttt{Int} or \texttt{Char}. Although, in theory, these two statements are to the same effect, programmers are often required to compute the latter from the former or vice versa. And this computation may carry out multiple layers. Programmers have to remember all the intermediate types and their reasoning throughout such mental gymnastics. Assisted by candidate expression cards and deduction steps, this intermediate information is externalized on screen and can be retrieved anytime. A recent study on working memory \cite{crichton_role_2021} suggested this approach may provide a positive effect in helping programmers manage cognitive load and free up working-memory space for high-level thinking.

\begin{listing}
\begin{minted}{haskell}
f z     
    | z == 3 = False
    | z == '4' = True
\end{minted}
\caption{In this simple program, \chameleon{} reports the error happened in \texttt{f} and \texttt{z}.}
\label{listing:2}
\end{listing}
% Implication for research
% In our studies, \chameleon{} shows a few different designs for type error visualization and interaction. However, we are far from exhausting the search space. One challenge we notice in the current design is that programmers have to shift their focus between the editor on the left and the \chameleon{} debugging interface on the right. Using the hover popup window in most mainstream IDEs may reduce the context switching during debugging.

% Integration with existing tools
% The current implementation of \chameleon{} requires non-trivial adaptation for editors such as VS Code and IntelliJ due to the overlay explanation layer. This type of error visualization is non-standard in mainstream editors. However, there may exist alternative representations of \chameleon{} error reporting using only the features available in major editors and IDEs. It is especially beneficial to represent \chameleon{} errors using a universal debugging middle layer such as language server protocol (LSP). This will allow \chameleon{}  to be adapted into various coding environments which support an LSP back end.

% Extend to other languages
% Our study focuses on the Haskell language for its popularity in the academic world. However, the low quality of textual type errors is a problem not limited to Haskell. Modern statically typed languages more or less all share the same problem. We believe the underlying type reporting technique and algorithms can be generalized to other languages. It will be exciting to see how interactive debugging features perform in other paradigms, such as  imperative languages and incremental type systems.

% Ranking candidate expressions and alternative types
% Our design of \chameleon{}  emphasizes allowing programmers to update the type error based on their prior knowledge. It unfolds to answer three questions: which expression should be considered ill-typed? Between multiple possible types, which one should be considered intended? Among all the possible locations, which one should be considered the cause? \chameleon{} answers all three questions by leaving them to the hands of the programmers. However, it is possible in situations, prescriptive error messages are preferred. Using heuristic methods may achieve the best of both worlds. It may produce effective results by guessing the possible un-typable expression based on the position it appears in the syntax tree and project structure or guessing the intended type signature by comparing the number of supporting slices in the source code.

### 2022-12-08 sections/evaluation.tex
More Tim's edits

Three studies were conducted to answer new research questions or rephrase existing ones. These studies are grouped into 2 phases, and each tested a different version of \chameleon{} with human participants. The number of participants and their Haskell experience, as well as key research questions and conclusions, were shown in figure \ref{fig:timeline}. \todo{Explain was a user study of the tool with practitioners?}

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{images/timeline.pdf}
    \caption{The timeline of \chameleon{}  evaluation.}
    \label{fig:timeline}
\end{figure*}

@@

The recruiting channel we used was the social news aggregation website Reddit. Specifically, our user study advertisements were posted in the Haskell programming language community \textit{r/haskell} and a general "programming language" focused community \textit{r/programminglanguages}. Recruiting from social media allowed us to access a more diverse demographic that better represent the true population of Haskell programmers. The participation is fully anonymized. The detailed ethical implications of these experiments are reviewed and approved by the IRB of the institution of the authors.

@@

\subsubsection*{\textbf{Experiment setting}}
The experiments took place remotely and unsupervised. Participants took the study online via a web browser and at the physical venue of their choosing. All user studies use a web-based debugging environment developed by the authors. Conducting the studies online helped us avoid variation when performing tasks in unfamiliar places and using different setups.

\subsubsection*{\textbf{Training and group assignment}}
After consent, participants received interactive training on the tool interface and interactive features. Participants were also shown a cheat sheet summarizing the key functionality of the interface. Participants had access to the cheat sheet at all times during the study. Participants were given 4 trial runs (2 for each setting) before the data collection started.

All the user studies used a within-subject design to evaluate the effectiveness of different tools or feature sets while counterbalancing the difference in programming proficiency between participants. In each study, participants were required to complete a series of programming tasks (8 for studies 1a and 1b, 9 for study 2). At each task, a participant receives a single Haskell file that contains one or more type errors. The participant was asked to make the code type check with the help of the given tool.

\subsubsection*{\textbf{Measurement}}
Time is measured from the start of each task to the first time the program is successfully type-checked and also passes all the functional tests. The data is automatically recorded by the online debugging environment. To not introduce a barrier to completing the study, every task can be skipped if the participant made three failed attempts or is stuck for over 1 minute on the task.

After completing all the tasks, participants are prompted to complete a debriefing survey. The survey questions include their Haskell experience and feedback on the tools and feature set participants used during the study.

We used a browser session recording tool~\cite{openreplay_openreplay_2022} to record the study sessions. This provides us with a "poor man's video recording" to identify usability issues in the study and to recognize general patterns. The reason we found this format inferior is that the recording technology does not provide a high enough sampling rate for us to be confidently used for rigorous analysis.

One use of the session recording is for outlier trimming.  We identify participants who left the computer for an extended period of time. We first ran a statistical analysis to identify the long gap (2 standard deviations) between actions using program logs for each participant. Then, we manually checked with the session recording to verify there was actually no user input (mouse movement, scrolling).

@@

\subsubsection{\textbf{\chameleon{} 1}}  
\chameleon{} 1 is an earlier iteration of \chameleon{} development. This version features the type inference engine that recovers most concrete types after type errors occur and a minimal set of debugging features. Key features in \chameleon{} 1 include showing two (or more) alternative types, showing all possible error locations, dividing possible error locations into groups based on alternative types, and concrete type restoration. In short, \chameleon{} 1 is equivalent to \chameleon{} 2 set to basic mode.

@@

Two  studies (1a \& 1b) were conducted to compare the effectiveness of solving type errors using \chameleon{} 1 and GHC compiler error messages. We choose GHC compiler error messages as the baseline because it is the canonical tool for working with type errors in Haskell. Although high-level tools like Haskell Language Server exist, they generally relay the GHC error messages verbatim. The procedure of these studies is shown in figure \ref{fig:procedure-1}.

@@

The  studies (1a \& 1b) are designed to answer the research question:
\begin{itemize}
    \item \textbf{RQ1.} Do programmers solve type errors faster with \chameleon{} than GHC compiler error messages?
\end{itemize}

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/procedure-1.pdf}
    \caption{\todo{Does this figure really help at all?}}
    \label{fig:procedure-1}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/user-study-1a.pdf}
    \caption{Time to complete each task in user study 1a with 95\% confidence interval}
    \label{fig:analysis-1a}
\end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/user-study-1b.pdf}
    \caption{Time to complete each task in user study 1b with 95\% confidence interval}
    \label{fig:analysis-1b}
\end{figure}

@@

From the data collected during study 1a (\ref{fig:analysis-1a}), we were not able to identify which group performed better. For the trivial challenges we set for users, the individual differences are generally more significant than differences between treatments. One interesting observation is task 8, where the \chameleon{} group outperformed the GHC group more consistently (p-value = 0.002456, Wilcoxin signed test). The discrepancy is thought to be related to the nature of Task 8. It has a lengthier source file and involves more language features (abstract data types and high-level functions). GHC struggles to produce a relevant error message for this type of error. We hypothesized that we may observe a more significant difference using tasks with lengthier and more realistic source code. This hypothesis is also supported by the most common feedback (7 out of 35) the study 1a was that the tasks were too trivial to invite meaningful evaluation. One participant said, "Looks nicer than GHC, but without trying it on something more complicated, I cannot conclude whether it would help me in practice."

In study 1b, the \chameleon{} group solved the type error faster (p =0.00742, Wilcoxin signed test) than the GHC group in almost all tasks (figure \ref{fig:analysis-1b}), barring task 1. For task 1, it is suspected that some participants spent more time exploring the interface of \chameleon{} due to its unfamiliarity and ineffective training. For all the other tasks, from the video recordings, we saw many \chameleon{} users confidently skip reading unrelated chunks of code, while GHC users generally read through the whole program. In harder problems and messier code, we notice programmers start to report the benefits of \chameleon{}. "It's most useful feature that I noticed was that it points out the locations of both conflicting uses; GHC often makes it difficult to figure out how it's coming to a conclusion about a type." reported one participant. "I think \chameleon{}  does a much better job than GHC's error messages. I like that it shows the sources for the type judgments. This makes it quite easy to figure out how to rectify errors." reported another participant.

@@

\subsubsection{\textbf{\chameleon{} 2}}  \label{sub:us4}

\chameleon{} 2 is latest version of \chameleon{}. Advanced debugging features such as deduction steps and candidate expressions were available in this iteration. In addition, mode switching is enabled by default as well. A few other user interfaces were designed and prototyped between the development of \chameleon{} 1 and \chameleon{} 2. Study 2 was designed to evaluate \chameleon{} 2.  We were trying to answer the research question: \textit{How do programmers use the interactive features in \chameleon{} 2?}. More specifically, we were trying to inquiry the following three sub-questions:
\begin{itemize}
    \item Do programmers prefer switching modes during debugging type errors?
    \item What are programmers' preferences among the three modes provided by \chameleon{} 2?
    \item How do programmers use the advanced features provided by \chameleon{} 2?

@@

During the study, the initial mode of each task alternated through the three different modes and repeated three cycles in nine tasks. The order of the three modes in each cycle is counterbalanced among all participants. However, participants can switch to other modes at any time (figure \ref{fig:procedure-2}).

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/procedure-2.pdf}
    \caption{\todo{Does this figure really help at all?}}
    \label{fig:procedure-2}
\end{figure}

@@

The study of \chameleon{} 2 is an exploratory study. We were expecting to let programmers discover their own way of using the tool. In post hoc analysis of the collected log data, we were able to extrapolate some interesting patterns of how the tool was used. We share a few of our observations in this section.

The most striking feature of the data is users' tendencies to use the tool differ wildly. Some users used the feature extensively however some users completed the tasks without actively exploring the given information. Based on this discrepancy, we divided the users into three groups.

\begin{tabularx}{\linewidth}{ 
  | >{\raggedright\arraybackslash}X 
  | >{\raggedright\arraybackslash}X  | }

    \hline
        Interaction level & Description \\ \hline
        Minimal Interaction & Users completed the tasks by making changes in source code, type checking, and reading error messages. \\ \hline
        Low Interaction & Users only actively used universal features in all modes, for example, hovering on "Possible type 1" and "Possible type 2" to narrow down error space. \\ \hline
        High Interaction & Users did everything from the Common feature group but used features specific to the Balanced mode and the Advanced mode, such as activating steps and expression cards. \\ \hline
\end{tabularx}

As shown in  figure \ref{fig:r4-analysis}, the time to complete each task roughly relates to the interaction level of participants. Participants with higher interaction levels generally performed better in the tasks, and the lowest level was worse (one-way ANOVA since we have three samples, p = 0.0368). The results from three tasks stand out from the general trend: in Tasks 4 and 6, higher interaction users performed worse, and in task 9, the general trend is exaggerated. As discussed in user studies 1 and 2, it is likely related to the nature of these tasks. Tasks 4 and 6 are shorter than other tasks. The ideal fixes for these two tasks are placed relatively early in the source code (both in the first two lines of the source code). These may be useful because following the natural reading order will allow participants to skip lines on the bottom, and thus the advantage of \chameleon{} is less critical. On the other hand, task 9 is the lengthiest task of all. It also involves more complex features, such as mutually recursive type definitions.

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/user-study-2.pdf}
    \caption{Time to complete each task in study 2 with 95\% confidence interval}
    \label{fig:r4-analysis}
\end{figure}

@@

Another observation is when using the mode switching feature of \chameleon{}, we find presenting the starting mode and finishing mode of each task and each participant in a confusion matrix is very telling (Fig. \ref{fig:r4-mode-switching}). 
This observation suggests two characteristics of using multi-mode debugging tools. First, programmers are generally reluctant to switch modes (146/277 changing modes vs. 131/277 staying in the same mode). Second, when changing modes, programmers generally switch to the more informative modes instead of, the more concise ones.
\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/mode-switching.pdf}
    \caption{
        % The rows represent the finishing mode, columns starting modes.
    }
    \label{fig:r4-mode-switching}
\end{figure}

>>>>

We conducted three user studies, iteratively refining the \chameleon{} UI and evaluating several research questions as per Fig.~\ref{fig:timeline}. 
%The number of participants and their Haskell experience, as well as key research questions and conclusions, were shown in figure \ref{fig:timeline}. \todo{Explain was a user study of the tool with practitioners?}

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth,trim=30mm 35mm 35mm 5mm]{images/timeline.pdf}
    \caption{The timeline of \chameleon{}  evaluation.}
    \label{fig:timeline}
\end{figure}

@@

Participants were recruited via the Reddit \textit{r/haskell} and \textit{r/programminglanguages} communities. 
%Recruiting from social media allowed us to access a more diverse demographic that better represent the true population of Haskell programmers. 
Participation is fully anonymized; detailed ethical implications of these experiments are reviewed and approved by the IRB of the institution of the authors.

@@

\subsubsection*{\textbf{Experiment setting}}
Experiments were conducted online and unsupervised. 
%Participants took the study online via a web browser and at the physical venue of their choosing. 
All user studies use a web-based debugging environment developed by the authors. 
%Conducting the studies online helped us avoid variation when performing tasks in unfamiliar places and using different setups.

\subsubsection*{\textbf{Training and group assignment}}
After consent, participants received interactive training on the tool interface and interactive features. Participants were also shown a cheat sheet summarizing the key functionality of the interface, and had access to the cheat sheet at all times during the study. Participants were given 4 trial runs (2 for each setting) before the data collection started. 
All the studies used a within-subject design to evaluate the effectiveness of different tools or feature sets while counterbalancing the difference in programming proficiency between participants. In each study, participants were required to complete a series of programming tasks (8 for studies 1a and 1b, 9 for study 2). At each task, a participant receives a single Haskell file that contains one or more type errors. The participant was asked to correct the code with the help of the given tool.

\subsubsection*{\textbf{Measurement}}
Time is measured from the start of each task to the first time the program is successfully type-checked and also passes all the functional tests. The data is automatically recorded by the online debugging environment. To not introduce a barrier to completing the study, every task can be skipped if the participant made three failed attempts or is stuck for over 1 minute on the task.
After completing all tasks, participants are prompted to complete a debriefing survey. The survey questions include their Haskell experience and feedback on the tools.

We used a browser session recording tool~\cite{openreplay_openreplay_2022} to record the study sessions. This provides us with a "poor man's video recording" to identify usability issues in the study and to recognize general patterns. 
%The reason we found this format inferior is that the recording technology does not provide a high enough sampling rate for us to be confidently used for rigorous analysis.
%One use of the session recording is for outlier trimming.  We identify participants who left the computer for an extended period of time. We first ran a statistical analysis to identify the long gap (2 standard deviations) between actions using program logs for each participant. Then, we manually checked with the session recording to verify there was actually no user input (mouse movement, scrolling).

@@

\subsubsection{\textbf{\chameleon{} 1}}  
An earlier version of the UI than that depicted in Figs.~(2-13), it featured the type inference engine that recovers most concrete types after type errors occur and a minimal set of debugging features. Key features in \chameleon{} 1 include showing two (or more) alternative types, showing all possible error locations, dividing possible error locations into groups based on alternative types, and concrete type restoration. In short, \chameleon{} 1 is equivalent to \chameleon{} 2 set to basic mode.

@@

Two  studies (1a \& 1b) were conducted to compare the effectiveness of solving type errors using \chameleon{} 1 and GHC compiler error messages. We choose GHC compiler error messages as the baseline because it is the canonical tool for working with type errors in Haskell.
%Although high-level tools like Haskell Language Server exist, they generally relay the GHC error messages verbatim. 
% The procedure of these studies is shown in figure \ref{fig:procedure-1}.

@@

Studies (1a \& 1b) address the research question:

\noindent\textbf{RQ1.} \textit{Do programmers solve type errors faster with \chameleon{} than GHC compiler error messages?}

% \begin{figure}[h]
%     \centering
%     \includegraphics[width=\linewidth]{images/procedure-1.pdf}
%     \caption{\todo{Does this figure really help at all?}}
%     \label{fig:procedure-1}
% \end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.8\linewidth,trim=0mm 12mm 0mm 35mm,clip]{images/user-study-1a.pdf}
    \caption{Study 1a task completion time (secs.) with 95\% confidence interval.}
    \label{fig:analysis-1a}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.8\linewidth,trim=0mm 15mm 0mm 35mm,clip]{images/user-study-1b.pdf}
    \caption{Study 1b task completion time (secs.) with 95\% confidence interval.}
    \label{fig:analysis-1b}
\end{figure}

@@

The data collected during study 1a, Fig.~\ref{fig:analysis-1a} does not show significant differences across Tasks 1-7. In hindsight, these tasks were trivial challenges for most users and the individual differences of participants are generally more significant than differences between treatments. However, one interesting observation is task 8, where the \chameleon{} group outperformed the GHC group (p-value = 0.002456, Wilcoxin signed test). We attribute this significant difference to the hardness of Task 8. The source file is longer and involves more language features (abstract data types and high-level functions). GHC struggles to produce a relevant error message for this type of error. 
From this result, we hypothesized that we may observe a more significant difference using tasks with lengthier and more realistic source code. 
This hypothesis is also supported by the most common feedback (7 out of 35) that the tasks were too trivial to invite meaningful evaluation. One participant said, ``Looks nicer than GHC, but without trying it on something more complicated, I cannot conclude whether it would help me in practice."

Therefore, in study 1b we introduced more difficult challenges and indeed observed that the \chameleon{} group were faster (p =0.00742, Wilcoxin signed test) than the GHC group in almost all tasks (figure \ref{fig:analysis-1b}), barring task 1. For task 1, it is suspected that some participants spent more time exploring the interface of \chameleon{} due to its unfamiliarity. For all other tasks, from the video recordings, we saw many \chameleon{} users confidently skip reading unrelated chunks of code, while GHC users generally read through the whole program. In harder problems and messier code, we notice programmers start to report the benefits of \chameleon{}. ``It's most useful feature that I noticed was that it points out the locations of both conflicting uses; GHC often makes it difficult to figure out how it's coming to a conclusion about a type." reported one participant. ``I think \chameleon{}  does a much better job than GHC's error messages. I like that it shows the sources for the type judgments. This makes it quite easy to figure out how to rectify errors." reported another participant.

@@

\subsubsection{\textbf{\chameleon{} 2}}  \label{sub:us4}
Based on observations of Study 1 we introduced several new features to \chameleon{}, eventually resulting in the UI depicted in Figs.~(2-13). Advanced debugging features such as deduction steps and candidate expressions were available in this iteration. In addition, mode switching is enabled by default. A few other user interfaces were designed and prototyped between the development of \chameleon{} 1 and \chameleon{} 2. Study 2 therefore addresses the research question:

\noindent\textbf{RQ2:} \textit{How do programmers use the interactive features in \chameleon{} 2?}.

More specifically:
\begin{itemize}
    \item Do programmers prefer switching modes during debugging type errors?
    \item What are programmers' preferences among the three modes provided by \chameleon{} 2?
    \item How do programmers use the advanced features provided by \chameleon{} 2?

@@

During the study, the initial mode of each task alternated through the three different modes and repeated three cycles in nine tasks. The order of the three modes in each cycle is counterbalanced among all participants. However, participants can switch to other modes at any time.

% \begin{figure}[h]
%     \centering
%     \includegraphics[width=\linewidth]{images/procedure-2.pdf}
%     \caption{\todo{Does this figure really help at all?}}
%     \label{fig:procedure-2}
% \end{figure}

@@

Study 2 is more exploratory in methodology than Study 1. 
We were expecting to let programmers discover their own way of using the tool. In post hoc analysis of the collected log data, we were able to extrapolate some interesting patterns of how the tool was used. %We share a few of our observations in this section.

The most striking feature of the data is that users tend to vary wildly in their use of the tool. Some users used the features extensively while others completed the tasks without actively exploring the given information. Based on this discrepancy, we divided the users into three groups:

\begin{small}
\noindent\begin{tabularx}{\linewidth}{ 
  | >{\hsize=.3\hsize}X 
  | >{\hsize=.7\hsize \raggedright\arraybackslash}X  | }

    \hline
        Interaction level & Description \\ \hline
        \textit{Minimal}  & Users completed the tasks by making changes in source code, type checking, and reading error messages. \\ \hline
        \textit{Low}  & Users only actively used universal features in all modes, for example, hovering on "Possible type 1" and "Possible type 2" to narrow down error space. \\ \hline
        \textit{High}  & Users did everything from the Common feature group but used features specific to the Balanced mode and the Advanced mode, such as activating steps and expression cards. \\ \hline
\end{tabularx}
\end{small}

As shown in  Fig.~\ref{fig:r4-analysis}, time to complete each task roughly relates to the interaction level of participants. Participants with higher interaction levels generally performed better, and the lowest interaction level was worse (one-way ANOVA since we have three samples, p = 0.0368). The results from three tasks stand out from the general trend: in Tasks 4 and 6, higher interaction users performed worse, and in task 9, the general trend is exaggerated. As with Study 1a and 1b, this difference is likely related to task difficulty. Tasks 4 and 6 are shorter than other tasks. The ideal fixes for these two tasks are placed relatively early in the source code (both in the first two lines of the source code). These may be useful because following the natural reading order will allow participants to skip lines on the bottom, and thus the advantage of \chameleon{} is less critical. On the other hand, task 9 is the lengthiest task of all. It also involves more complex features, such as mutually recursive type definitions.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth,trim=0mm 15mm 0mm 35mm,clip]{images/user-study-2.pdf}
    \caption{Study 2 task completion time (secs.) with 95\% confidence intervals.}
    \label{fig:r4-analysis}
\end{figure}

@@

Another observation is when using the mode switching feature of \chameleon{}, we find presenting the starting mode and finishing mode of each task and each participant in a confusion matrix is very telling (Fig. \ref{fig:r4-mode-switching}). 
This observation suggests two characteristics of using multi-mode debugging tools. First, programmers are generally reluctant to switch modes (146/277 changing modes vs. 131/277 staying in the same mode). Second, when changing modes, programmers generally switch to the more informative modes instead of, the more concise ones.
\begin{figure}
    \centering
    \includegraphics[width=0.7\linewidth,trim=0mm 5mm 0mm 5mm,clip]{images/mode-switching.pdf}
    \caption{Study 2 mode switches by starting mode.  Users overwhelmingly switched to the more sophisticated interface mode.
        % The rows represent the finishing mode, columns starting modes.
    }
    \label{fig:r4-mode-switching}
\end{figure}

\subsection{Limitations}

One threat to the validity of the evaluation is the number of participants. Although for each study we received hundreds of online participants, the studies suffer from a high abandonment rate (especially study 1b). This was expected: the programming challenges are difficult and our volunteer participants are unremunerated. 
%To encourage completion, we included a cheat sheet containing a summary of the key functionality of the tools and necessary type signatures. A skip button is available for when participants are stuck.
Because we recruited participants online and anonymized all the participants, it is possible for participants of a previous study to enter a later one. This creates variation in familiarity. We offset this by using new code challenges in every study and conducting trial runs before data collection to bring new participants up to speed.
Conducting studies remotely and unsupervised there is no way to intervene when users encounter usability issues. To mitigate this we conducted cognitive walkthroughs and sandbox pilots before running each study.

Future evaluation would benefit from using more realistic tasks. In studies 1b and 2, we used more challenging tasks compared to study 1a. However, the complexity is just short of what professional Haskell programmers may face in the production Haskell codebase. It would be interesting to see how \chameleon{} is used against type errors that span multiple files and packages and include confusing abstractions, like Monads, Monad transformers, and Lenses.

% The online study limited the amount of qualitative data we could gather from the users. Although we handed out debriefing surveys after each study, it is no substitute for more in-depth qualitative study methods. Inviting users for speak-aloud walkthroughs and interviews are viable methods to unveil deeper usability issues about \chameleon{}.
